The Unicode Cookbook \linebreak for Linguists

### 2017-02-02 book/chapters/introduction.tex

Before we deal with these proposals we will first discuss the theoretical
background on text encoding, on the Unicode standard, and on the International
Phonetic Alphabet. In the remainder of this chapter, we give an extended
introduction to the notion of encoding (Section~\ref{encoding}) and the
principles of writing systems, both from a linguistic perspective
(Section~\ref{linguistic-terminology}) and from the perspective of the Unicode
Consortium (Section~\ref{the-unicode-approach}). We consider the Unicode
Standard to be a breakthrough (and ongoing) development that fundamentally
changed the way we look at writing systems, and we aim to provide here a
slightly more in-depth survey of the many techniques that are available in the
standard. A good appreciation for the solutions that the Unicode Standard
provides, also allows for a thorough understanding of possible pitfalls that one
might encounter when using it (Chapter~\ref{unicode-pitfalls}). As an example of
the current state-of-the-art, we discuss the International Phonetic Alphabet
(IPA, Chapter~\ref{the-international-phonetic-alphabet}) and its rather
problematic marriage with the Unicode Standard (Chapter
\ref{ipa-meets-unicode}).

The second part of the book describes our proposals for how to deal with the
Unicode Standard in the daily practice of linguistics. First, we discuss the
challenges of characterizing a writing system. To solve these problems, we
propose the notions of orthography profiles, closely related to Unicode locale
descriptions (Chapter~\ref{orthography-profiles}). We provide reference
implementation of our proposals in R and in Python, available as open-source
libraries. An introduction to these implementations of our proposals is
described in Chapter~\ref{implementation}. Finally, we discuss practical issues
with actual examples in Chapter~\ref{use-cases}.

The following conventions are followed in this paper. All phonemic and phonetic
representations are given in the International Phonetic Alphabet (IPA), unless
noted otherwise \citep{IPA2005}. Standard conventions are used for
distinguishing between graphemic < >, phonemic / / and phonetic [ ]
representations. For character descriptions, we follow the notational
conventions of the Unicode Standard \citep{Unicode2014}. Character names are
represented in small capital letters (e.g.~\textsc{latin small letter schwa})
and code points are expressed as U\emph{+n} where \emph{n} is a four to six
digit hexadecimal number, e.g.~\uni{0256}, which can be rendered as the glyph <ə>.

>>>>

Before we deal with these proposals we will first discuss the theoretical
background on text encoding, on the Unicode standard, and on the International
Phonetic Alphabet. In the remainder of this chapter, we give an extended
introduction to the notion of encoding (Section~\ref{encoding}) and the
principles of writing systems, both from a linguistic perspective
(Section~\ref{linguistic-terminology}) and from the perspective of the Unicode
Consortium (Section~\ref{the-unicode-approach}). We consider the Unicode
Standard to be a breakthrough (and ongoing) development that fundamentally
changed the way we look at writing systems, and we aim to provide here a
slightly more in-depth survey of the many techniques that are available in the
standard. A good appreciation for the solutions that the Unicode Standard also
allows for a thorough understanding of possible pitfalls that one might
encounter when using it in general (Chapter~\ref{unicode-pitfalls}). Linguists are more often interested in using the Unicode Standard with the International Phonetic Alphabet (IPA). We first provide a history of the development of the IPA and early attempts to encode it electronically (Chapter~\ref{the-international-phonetic-alphabet}) before we discuss the rather problematic marriage of the International Phonetic Alphabet (IPA) with the Unicode Standard (Chapter\ref{ipa-meets-unicode}).

In the second part of the book (Chapters~\ref{orthography-profiles} and
\ref{implementation}) we describe our proposals of how to deal with the Unicode
Standard in the daily practice of (comparative) linguistics. First, we discuss the
challenges of characterizing a writing system. To solve these problems, we
propose the notions of orthography profiles, closely related to Unicode locale
descriptions (Chapter~\ref{orthography-profiles}). Finally, we discuss practical
issues with actual examples (Chapter~\ref{implementation}). We provide reference
implementation of our proposals in R and in Python, available as open-source
libraries. An introduction to these implementations of our proposals is
described in Chapter~\ref{implementation}. Finally, we discuss practical issues
with actual examples in Chapter~\ref{use-cases}.

\subsubsection*{Conventions}

The following conventions are followed in this book. All phonemic and phonetic
representations are given in the International Phonetic Alphabet (IPA), unless
noted otherwise \citep{IPA2005}. Standard conventions are used for
distinguishing between graphemic < >, phonemic / / and phonetic [ ]
representations. For character descriptions, we follow the notational
conventions of the Unicode Standard \citep{Unicode2014}. Character names are
represented in small capital letters (e.g.~\textsc{latin small letter schwa})
and code points are expressed as U\emph{+n} where \emph{n} is a four to six
digit hexadecimal number, e.g.~\uni{0256}, which can be rendered as the glyph <ə>.

### 2017-02-02 book/chapters/ipa_meets_unicode.tex
minor changes

Conversely, non-linguists are unlikely to distinguish any semantic difference
between an open back unrounded vowel, which is encoded in IPA with a
`single-story' <ɑ> \textsc{latin small letter alpha} at \uni{0251}, and the open
front unrounded vowel, which is encoded in IPA with a `double-story' <a>
\textsc{latin small letter a} at \uni{0061}, basically treating them as
homoglyphs, although they are different phonetic symbols. But even among
linguists this distinction leads to problems. For example, as pointed out by
\citet{Mielke2009}, there is a problem stemming from the fact that about 75\% of
languages are reported to have a five-vowel system \citep{Maddieson1984}.
Historically, linguistic descriptions tend not to include precise audio
recording and measurements of formants, so this may lead one to ask if the many
<a> characters that are used in phonological description reflects a
transcriptional bias. The common use of <a> in transcriptions could be in part
due to the ease of typing the letter on an English keyboard (or for older
descriptions, the typewriter). We found it to be exceedingly rare that a
linguist uses <ɑ> for a low back unrounded vowel.\footnote{One example is
\citet[75]{Vidal2001a}, in which the author states: ``The definition of Pilagá
/a/ as [+back] results from its behavior in certain phonological contexts. For
instance, uvular and pharyngeal consonants only occur around /a/ and /o/. Hence,
the characterization of /a/ and /o/ as a natural class of (i.e., [+back]
vowels), as opposed to /i/ and /e/.''} They simply use <a> as long as there is
no opposition to <ɑ>.

@@

% ==========================
\section{Pitfall: Future revisions to the IPA}
\label{ipa-revisions}
% ==========================

>>>>

Conversely, non-linguists are unlikely to distinguish any semantic difference
between an open back unrounded vowel, which is encoded in IPA with a
`single-story' <ɑ> \textsc{latin small letter alpha} at \uni{0251}, and the open
front unrounded vowel, which is encoded in IPA with a `double-story' <a>
\textsc{latin small letter a} at \uni{0061}, basically treating them as
homoglyphs, although they are different phonetic symbols. But even among
linguists this distinction leads to problems. For example, as pointed out by
\citet{Mielke2009}, there is a problem stemming from the fact that about 75\% of
languages are reported to have a five-vowel system \citep{Maddieson1984}.
Historically, linguistic descriptions tend not to include precise audio
recording and measurements of formants, so this may lead one to ask if the many
<a> characters that are used in phonological description reflects a
transcriptional bias. The common use of <a> in transcriptions could be in part
due to the ease of typing the letter on an English keyboard (or for older
descriptions, the typewriter). We found it to be exceedingly rare that a
linguist uses <ɑ> for a low back unrounded vowel.\footnote{One example is
\citet[75]{Vidal2001}, in which the author states: ``The definition of Pilagá
/a/ as [+back] results from its behavior in certain phonological contexts. For
instance, uvular and pharyngeal consonants only occur around /a/ and /o/. Hence,
the characterization of /a/ and /o/ as a natural class of (i.e., [+back]
vowels), as opposed to /i/ and /e/.''} They simply use <a> as long as there is
no opposition to <ɑ>.

@@

% ==========================
\section{Pitfall: Revisions to the IPA}
\label{ipa-revisions}
% ==========================

### 2017-02-02 book/chapters/preface.tex

% ==========================
% \section{Preface}
% \label{preface}
% ==========================

% Early phoneticians: Richard Lepsius, Isaac Pitman, Alexander J. Ellis, and Alexander Melville Bell
% Later ones: Henry Sweet, L'abbé Jean-Pierre Rousselot, Eduard Sievers, Paul Passy, and Jens Otto Harry Jesperson

In the 1840s, several researchers and teachers began developing a notation to accurately transcribe articulatory gestures and their acoustic correlates. Their goal was not only transcription, they also wanted to use their notation for spelling reform (especially English), for teaching literacy and foreign languages, and in particular for developing writing systems for unwritten languages.

At first, the challenge of devising a universal phonetic notation seemed straightforward: normalize the relationship between graphemes and sounds (<c> in `cake' -> /k/; <c> in `cell' -> /s/) to remove inconsistencies introduced by the historical divergence of written and spoken forms (e.g.\ in English and French orthographies), and keep adding new symbol-sound mappings until all sounds are accounted for.

% These early transcription system pioneers found success with the first few languages that they encountered. Grapheme replacement and normalization of symbols across different orthographies seemed to work. But as they were confronted with more articulations and different notational devices, early phoneticians encountered a wealth of potentially important data that they had not expected.

This approach was successful for the first few languages investigated. But as the early phoneticians were confronted with more and more articulations, they encountered a wealth of data that they had not expected. The in-depth exploration of articulatory phonetics led to the discovery that the number of possible spoken sounds is infinite. \cite{Hockett1995} notes:

\begin{quote}
``But as they observed articulation more and more closely, and as they took more and more languages into consideration, the number of sounds they could discriminate and the number of symbols needed to denote them both increased seemingly without limit. In fact, as early as 1848, Ellis asserted that in principle the number of different sounds is endless. With the development of instrumental phonetics, largely by L'abbe Rousselt, it soon turned out that, given sufficiently accurate measuring devices, that is in fact the case.''
\end{quote}

% Hockett 1995: "And so the phoneticians began to develop notations that would show only the constant and distinctive features and let the transient and predictable ones be provided for by cover statements"

% The phonemic principle emerged and the model for it was alphabetic writing.

The first instrumental measurements in the 1870-80s showed unequivocally that spoken language is comprised of sounds waves; it is dynamic and not discrete. Given the fluid nature of the speech stream, sounds differ from one utterance to the next and the same sound may differ in different environments. However, many of these changes are predictable in the context of those environments and they need not necessarily form distinctive lexical contrasts.

These findings left the early phoneticians in a conundrum given the aims of developing precise phonetic notion, but also practical alphabetic-based writing. It is in this vein that the phonetic and phonemic distinction was introduced very early into transcription. The etic vs emic principle took hold, in essence, even before terminology for the distinction was introduced and adopted.

A transcription system that showed both contrastive and non-contrastive elements emerged, taking form notably in the International Phonetic Alphabet (which itself has been updated from time-to-time over the last nearly 120 years). The association between phonetic and phonemic would be maintained in the International Phonetic Alphabet and through the various instantiations of phonological theory, which developed from the work of the early phoneticians.

The model for the principle of contrast was alphabetic writing, in particular shallow orthography, i.e.\ a one-to-one mapping between sounds and symbol. Traditional alphabetic orthography, however, did not scale to precise phonetic notation, so the former was sprinkled with diacritics. This dichotomy  causes many issues -- linguistic, theoretic, and recently technological -- which we bring to light in this book.

>>>>

This text is meant as a practical guide for linguists, and programmers, who work with data in multilingual computational environments. We introduce the basic concepts needed to understand how writing systems and character encodings function, and how they work together.

The intersection of the Unicode Standard and the International Phonetic Alphabet is often not met without frustration by users. Nevertheless, the two standards have provided language researchers with a consistent computational architecture needed to process, publish and analyze data from many different languages. We bring to light common, but not always transparent, pitfalls that researchers face when working with Unicode and IPA.

Our research uses quantitative methods to compare languages and uncover and clarify their phylogenetic relations. However, the majority of lexical data available from the world's languages is in author- or document-specific orthographies. Having identified and overcome the pitfalls involved in making writing systems and character encodings syntactically and semantically interoperable (to the extent that they can be), we created a suite of open-source Python and R tools to work with languages using profiles that adequately describe their orthographic conventions. Using orthography profiles and these tools allows users to segment text, analyze it, identify errors, and to transform it into different written forms.

We welcome comments and corrections of this text, our source code, and the use case studies that are supplement to this book. Please use the issue tracker or email us directly.

\bigbreak
\noindent Steven Moran {\textless}bambooforest@gmail.com{\textgreater}

\noindent Michael Cysouw {\textless}cysouw@mac.com{\textgreater}
